---
title: 'Reproducible Research: Peer Assessment No. 1'
author: "Bob Wothe"
date: "Friday, March 13, 2015"
output:
  html_document:
    self_contained: no
---
##LOADING AND PREPROCESSING THE DATA

This code will load the data and also slightly transform the data for easier processing.

```{r}
unzip("repdata-data-activity.zip")
dat <- read.csv("activity.csv")

dates <- strptime(dat$date, "%Y-%m-%d")
dat$date <- dates

uniqueDates <- unique(dates)
uniqueIntervals <- unique(dat$interval)
```

##WHAT IS THE MEAN TOTAL NUMBER OF STEPS PER DAY?

1. This code will create a histogram of the total number of steps taken each day. 

```{r, fig.width=11, fig.height=6}
stepsSplit <- split(dat$steps, dates$yday)
totalStepsPerDay <- sapply(stepsSplit, sum, na.rm=TRUE)
plot(uniqueDates, totalStepsPerDay, main="Steps taken each day", 
     xlab="Date (10/2012-11/2012)", ylab="Frequency", type="h", lwd=4, col="blue")
```

2.a. This code will calculate mean steps per day: 

```{r}
meanStepsPerDay <- sapply(stepsSplit, mean, na.rm=TRUE)
meanDataFrame <- data.frame(date=uniqueDates, meanStepsPerDay=meanStepsPerDay, row.names=NULL)
meanDataFrame
```

2.b. This code will calculate median steps per day:
```{r cache=TRUE}
medianStepsPerDay <- sapply(stepsSplit, median, na.rm=TRUE)
medianDataFrame <- data.frame(date=uniqueDates, medianStepsPerDay=medianStepsPerDay, row.names=NULL)
medianDataFrame
```

##WHAT IS THE AVERAGE DAILY ACTIVITY PATTERN?

1. This code will create a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, fig.width=11, fig.height=6}
intervalSplit <- split(dat$steps, dat$interval)
averageStepsPerInterval <- sapply(intervalSplit, mean, na.rm=TRUE)
plot(uniqueIntervals, averageStepsPerInterval, type="l",
     main="Average number of steps per interval across all days", 
     xlab="Interval", ylab="Average # of steps across all days", 
     lwd=2, col="blue")
maxIntervalDays <- max(averageStepsPerInterval, na.rm=TRUE)
maxIndex <- as.numeric(which(averageStepsPerInterval == maxIntervalDays))
maxInterval <- uniqueIntervals[maxIndex]
abline(v=maxInterval, col="red", lwd=3)
```

2. This codw will return the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps:

```{r cache=TRUE}
maxInterval
```

##INPUTTING MISSING VALUES

1. This code will calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r cache=TRUE}
completeRowsBool <- complete.cases(dat$steps)
numNA <- sum(as.numeric(!completeRowsBool))
numNA
```

2/3. This code will fill in all of the missing values in the dataset using the mean for that day and create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r cache=TRUE}
meanStepsPerDay[is.nan(meanStepsPerDay)] <- 0

meanColumn <- rep(meanStepsPerDay, 288)

rawSteps <- dat$steps

stepsNA <- is.na(rawSteps)

rawSteps[stepsNA] <- meanColumn[stepsNA]

datNew <- dat
datNew$steps <- rawSteps
```

This histogram will show the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r cache=TRUE, fig.width=11, fig.height=12}

stepsSplitNew <- split(datNew$steps, dates$yday)

totalStepsPerDayNew <- sapply(stepsSplitNew, sum)

par(mfcol=c(2,1))

plot(uniqueDates, totalStepsPerDayNew, main="Histogram of steps taken each day", 
     xlab="Date (10/2012-11/2012)", ylab="Frequency", type="h", lwd=4, col="blue")
```

Mean steps:

```{r cache=TRUE}
meanStepsPerDayNew <- sapply(stepsSplitNew, mean)
meanDataFrameNew <- data.frame(date=uniqueDates, meanStepsPerDay=meanStepsPerDay, 
                               meanStepsPerDayNew=meanStepsPerDayNew, row.names=NULL)
meanDataFrameNew
```

Median steps:

```{r cache=TRUE}
medianStepsPerDayNew <- sapply(stepsSplitNew, median)
medianDataFrameNew <- data.frame(date=uniqueDates, medianStepsPerDay=medianStepsPerDay, 
                                 medianStepsPerDayNew=medianStepsPerDayNew, row.names=NULL)
medianDataFrameNew
```

The impact of the missing data is minimal; this is likely due to an already representative sample being established.

##ARE THERE DIFFERENCES IN ACTIVITY PATTERNS BETWEEN WEEKDAYS AND WEEKENDS?

1. This code will create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r cache=TRUE}
wdays <- dates$wday

classifywday <- rep(0, 17568)

classifywday[wdays >= 1 & wdays <= 5] <- 1

classifywday[wdays == 6 | wdays == 0] <- 2

daysFactor <- factor(classifywday, levels=c(1,2), labels=c("Weekdays", "Weekends"))

datNew$typeOfDay <- daysFactor

datWeekdays <- datNew[datNew$typeOfDay == "Weekdays", ]
datWeekends <- datNew[datNew$typeOfDay == "Weekends", ]
```

2. This code will create two panel plots containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days and weekend days (y-axis). 

```{r cache=TRUE, fig.width=11, fig.height=12}
datSplitWeekdays <- split(datWeekdays$steps, datWeekdays$interval)
datSplitWeekends <- split(datWeekends$steps, datWeekends$interval)

meanStepsPerWeekdayInterval <- sapply(datSplitWeekdays, mean)
meanStepsPerWeekendInterval <- sapply(datSplitWeekends, mean)

par(mfcol=c(2,1))
plot(uniqueIntervals, meanStepsPerWeekdayInterval, type="l",
     main="Average number of steps per interval - weekdays", 
     xlab="Interval", ylab="Average # of steps - weekdays", 
     lwd=2, col="blue")
plot(uniqueIntervals, meanStepsPerWeekendInterval, type="l",
     main="Average number of steps per interval - weekends", 
     xlab="Interval", ylab="Average # of steps - weekends", 
     lwd=2, col="blue")
```